---
title: "03_statistical analysis"
output: html_document
date: "2026-07-15"
editor_options: 
  chunk_output_type: console
---
## Purpose

This script reproduces the statistical analyses, species summaries, and manuscript figures using the deposited vegetation and imputed trait datasets. Phylogeny construction and trait imputation are documented separately in `01_phylogeny_construction.Rmd` and `02_trait_imputation.Rmd`.

```{r setup}

library(readr)
library(tidyverse)
library(ade4)
library(scatterplot3d)
library(ggpubr)

```

## Read in datasets

```{r read-data}

traits_scaled <- read_csv(
  "traits_scaled.csv"
) %>%
  mutate(
    species = str_replace_all(
      species, "_", " ")
  )

traits_unscaled <- read_csv(
  "traits_unscaled.csv"
)

vegetation <- read_csv(
  "vegetation_data.csv",
  na = ""
)

```

## Join imputed traits to vegetation data

```{r join-data}

#function to prepare data for analysis
prepare_analysis_data <- function(vegetation_data, trait_data) {

  vegetation_data %>%
    mutate(
      category = case_when(
        family == "Asteraceae" & category == "exotic" ~ "Aster_exotic",
        family == "Asteraceae" & category == "native" ~ "Aster_native",
        TRUE ~ category
      ),
      category = factor(
        category,
        levels = c(
          "native",
          "Aster_exotic",
          "Aster_native"
        )
      )
    ) %>%
    filter(
      !is.na(category),
      !is.na(topography)
    ) %>%
    left_join(
      trait_data,
      by = "species"
    ) %>%
    mutate(
      presence = 1
    ) %>%
    select(
      topography,
      family,
      species,
      category,
      presence,
      height_m,
      SLA_mm2_mg.1,
      seed_dry_mass_mg
    ) %>%
    distinct() %>%
    pivot_wider(
      names_from = topography,
      values_from = presence,
      values_fill = 0
    ) %>%
    filter(
      !is.na(height_m),
      !is.na(SLA_mm2_mg.1),
      !is.na(seed_dry_mass_mg)
    )
}

analysis_scaled <- prepare_analysis_data(
  vegetation_data = vegetation,
  trait_data = traits_scaled
)

analysis_unscaled <- prepare_analysis_data(
  vegetation_data = vegetation,
  trait_data = traits_unscaled
)

dim(analysis_scaled)

dim(analysis_unscaled)

```

## Separate topographic groups

```{r topo-groups}

# separate topographic groups for scaled trait data
upper_slope <- analysis_scaled %>%
  filter(upper_slope == 1)

lower_slope <- analysis_scaled %>%
  filter(lower_slope == 1)

ridge_summit <- analysis_scaled %>%
  filter(ridge_summit == 1)

valley_flatland <- analysis_scaled %>%
  filter(valley_flatland == 1)

# and separate topographic groups for unscaled trait data
upper_slope_unscaled <- analysis_unscaled %>%
  filter(upper_slope == 1)

lower_slope_unscaled <- analysis_unscaled %>%
  filter(lower_slope == 1)

ridge_summit_unscaled <- analysis_unscaled %>%
  filter(ridge_summit == 1)

valley_flatland_unscaled <- analysis_unscaled %>%
  filter(valley_flatland == 1)

# check counts for species categories in each topo group
table(upper_slope$category)
table(lower_slope$category)
table(ridge_summit$category)
table(valley_flatland$category)


```

## Cumulative distribution function analyses

CDFs were calculated from species’ Euclidean distances to the centroid of native non-Asteraceae species in standardized three-dimensional trait space. Randomization tests compared native Asteraceae and exotic Asteraceae with the native reference group within each topographic group.

```{r cdf-functions}

calculate_cdf <- function(threshold, distances) {
  length(
    distances[abs(distances) <= threshold]
  ) / length(distances)
}

randomize_cdf <- function(
  scores,
  status,
  test_group,
  n_randomizations = 999
) {
  
  native_centroid <- apply(
    scores[status == "native", ],
    2,
    mean
  )
  
  n_test_species <- sum(
    status == test_group
  )
  
  distances <- as.matrix(
    dist(
      rbind(
        native_centroid,
        scores
      ),
      method = "euclidean"
    )
  )[, 1]
  
  distances <- distances[-1]
  
  x_values <- seq(
    from = 0,
    to = max(distances),
    length.out = 1000
  )
  
  pooled_distances <- distances[
    status %in% c(
      "native",
      test_group
    )
  ]
  
  randomized_cdfs <- vector(
    "list",
    n_randomizations
  )
  
  for (i in seq_len(n_randomizations)) {
    
    sampled_distances <- sample(
      pooled_distances,
      size = n_test_species,
      replace = FALSE
    )
    
    randomized_cdfs[[i]] <- vapply(
      x_values,
      calculate_cdf,
      FUN.VALUE = numeric(1),
      distances = sampled_distances
    )
  }
  
  randomized_cdfs
}

```

## Calculate observed and randomized CDFs

```{r calculate-cdfs}

topographic_data <- list(
  upper_slope = upper_slope,
  lower_slope = lower_slope,
  ridge_summit = ridge_summit,
  valley_flatland = valley_flatland
)

trait_columns <- c(
  "height_m",
  "SLA_mm2_mg.1",
  "seed_dry_mass_mg"
)

cdf_results <- vector(
  "list",
  length(topographic_data)
)

names(cdf_results) <- names(topographic_data)

for (topography_name in names(topographic_data)) {
  
  current_data <- topographic_data[[topography_name]]
  
  scores <- as.data.frame(
    current_data[, trait_columns]
  )
  
  status <- current_data$category
  
  native_centroid <- apply(
    scores[status == "native", ],
    2,
    mean
  )
  
  distances <- as.matrix(
    dist(
      rbind(
        native_centroid,
        scores
      )
    )
  )[, 1]
  
  distances <- distances[-1]
  
  x_values <- seq(
    from = 0,
    to = max(distances),
    length.out = 1000
  )
  
  observed_cdfs <- list(
    native = vapply(
      x_values,
      calculate_cdf,
      FUN.VALUE = numeric(1),
      distances = distances[status == "native"]
    ),
    
    Aster_native = vapply(
      x_values,
      calculate_cdf,
      FUN.VALUE = numeric(1),
      distances = distances[status == "Aster_native"]
    ),
    
    Aster_exotic = vapply(
      x_values,
      calculate_cdf,
      FUN.VALUE = numeric(1),
      distances = distances[status == "Aster_exotic"]
    )
  )
  
  set.seed(1234)
  
  randomized_native_asteraceae <- randomize_cdf(
    scores = scores,
    status = status,
    test_group = "Aster_native",
    n_randomizations = 999
  )
  
  set.seed(1234)
  
  randomized_exotic_asteraceae <- randomize_cdf(
    scores = scores,
    status = status,
    test_group = "Aster_exotic",
    n_randomizations = 999
  )
  
  cdf_results[[topography_name]] <- list(
    x = x_values,
    observed = observed_cdfs,
    randomized = list(
      Aster_native = randomized_native_asteraceae,
      Aster_exotic = randomized_exotic_asteraceae
    )
  )
}

```

## Test differences among CDFs

```{r cdf-test-functions}

area_between_cdfs <- function(
  cdf_1,
  cdf_2,
  x_values
) {
  
  differences <- (
    cdf_2 - cdf_1
  )[-1] * diff(x_values)
  
  negative_area <- sum(
    abs(
      differences[differences < 0]
    )
  )
  
  positive_area <- sum(
    differences[differences > 0]
  )
  
  signed_area <- if (
    negative_area > positive_area
  ) {
    -negative_area
  } else {
    positive_area
  }
  
  direction <- c(
    positive_area,
    negative_area
  ) / (
    positive_area +
      abs(negative_area)
  )
  
  list(
    signed_area,
    direction
  )
}

randomized_cdf_areas <- function(
  observed_cdf,
  randomized_cdfs,
  x_values
) {
  
  vapply(
    randomized_cdfs,
    function(randomized_cdf) {
      area_between_cdfs(
        cdf_1 = observed_cdf,
        cdf_2 = randomized_cdf,
        x_values = x_values
      )[[1]]
    },
    FUN.VALUE = numeric(1)
  )
}

```

```{r cdf-randomization-tests}

topography_labels <- c(
  upper_slope = "Upper Slope",
  lower_slope = "Lower Slope",
  ridge_summit = "Ridge + Summit",
  valley_flatland = "Valley + Flatland"
)

native_asteraceae_cdf_tests <- data.frame(
  topography = unname(topography_labels),
  dCDF = NA_real_,
  P = NA_real_
)

exotic_asteraceae_cdf_tests <- data.frame(
  topography = unname(topography_labels),
  dCDF = NA_real_,
  P = NA_real_
)

for (i in seq_along(cdf_results)) {
  
  current_result <- cdf_results[[i]]
  
  normalized_x <- current_result$x /
    max(current_result$x)
  
  native_asteraceae_simulated <- randomized_cdf_areas(
    observed_cdf = current_result$observed$native,
    randomized_cdfs =
      current_result$randomized$Aster_native,
    x_values = normalized_x
  )
  
  native_asteraceae_observed <- area_between_cdfs(
    cdf_1 = current_result$observed$native,
    cdf_2 = current_result$observed$Aster_native,
    x_values = normalized_x
  )[[1]]
  
  native_asteraceae_cdf_tests$dCDF[i] <-
    native_asteraceae_observed
  
  native_asteraceae_cdf_tests$P[i] <-
    as.randtest(
      sim = native_asteraceae_simulated,
      obs = native_asteraceae_observed,
      alter = "two-sided"
    )$pvalue
  
  
  exotic_asteraceae_simulated <- randomized_cdf_areas(
    observed_cdf = current_result$observed$native,
    randomized_cdfs =
      current_result$randomized$Aster_exotic,
    x_values = normalized_x
  )
  
  exotic_asteraceae_observed <- area_between_cdfs(
    cdf_1 = current_result$observed$native,
    cdf_2 = current_result$observed$Aster_exotic,
    x_values = normalized_x
  )[[1]]
  
  exotic_asteraceae_cdf_tests$dCDF[i] <-
    exotic_asteraceae_observed
  
  exotic_asteraceae_cdf_tests$P[i] <-
    as.randtest(
      sim = exotic_asteraceae_simulated,
      obs = exotic_asteraceae_observed,
      alter = "two-sided"
    )$pvalue
}

native_asteraceae_cdf_tests <- native_asteraceae_cdf_tests %>%
  mutate(
    Padj = p.adjust(
      P,
      method = "BH"
    )
  )

exotic_asteraceae_cdf_tests <- exotic_asteraceae_cdf_tests %>%
  mutate(
    Padj = p.adjust(
      P,
      method = "BH"
    )
  )

native_asteraceae_cdf_tests %>%
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 3)
    )
  )

exotic_asteraceae_cdf_tests %>%
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 3)
    )
  )

```

## Single trait randomization analyses

Differences in median trait values were tested separately within each topographic group. Native Asteraceae and exotic Asteraceae were each compared with native non-Asteraceae using 999 random permutations of species-group labels.

```{r single-trait-functions}

difference_of_medians <- function(
  trait_values,
  categories,
  test_group
) {
  
  data <- data.frame(
    trait = trait_values,
    category = categories
  ) %>%
    filter(
      complete.cases(.)
    )
  
  native_median <- median(
    data$trait[
      data$category == "native"
    ]
  )
  
  test_group_median <- median(
    data$trait[
      data$category == test_group
    ]
  )
  
  test_group_median - native_median
}

randomize_difference_of_medians <- function(
  trait_values,
  categories,
  test_group,
  n_randomizations = 999
) {
  
  data <- data.frame(
    trait = trait_values,
    category = categories
  ) %>%
    filter(
      complete.cases(.)
    )
  
  simulated_differences <- numeric(
    n_randomizations
  )
  
  for (i in seq_len(n_randomizations)) {
    
    randomized_categories <- sample(
      data$category
    )
    
    simulated_differences[i] <-
      difference_of_medians(
        trait_values = data$trait,
        categories = randomized_categories,
        test_group = test_group
      )
  }
  
  simulated_differences
}

```

```{r single-trait-tests}

topographic_data <- list(
  upper_slope = upper_slope,
  lower_slope = lower_slope,
  ridge_summit = ridge_summit,
  valley_flatland = valley_flatland
)

topography_labels <- c(
  upper_slope = "Upper Slope",
  lower_slope = "Lower Slope",
  ridge_summit = "Ridge + Summit",
  valley_flatland = "Valley + Flatland"
)

trait_names <- c(
  SLA = "SLA_mm2_mg.1",
  height = "height_m",
  seed_mass = "seed_dry_mass_mg"
)

run_single_trait_tests <- function(
  topographic_data,
  test_group
) {
  
  results <- expand_grid(
    topography = names(topographic_data),
    trait = names(trait_names)
  ) %>%
    mutate(
      observed_difference = NA_real_,
      P = NA_real_
    )
  
  for (i in seq_len(nrow(results))) {
    
    topography_name <- results$topography[i]
    trait_name <- results$trait[i]
    
    current_data <- topographic_data[[topography_name]] %>%
      filter(
        category %in% c(
          "native",
          test_group
        )
      )
    
    trait_column <- trait_names[[trait_name]]
    
    set.seed(1234)
    
    simulated_differences <-
      randomize_difference_of_medians(
        trait_values =
          current_data[[trait_column]],
        categories =
          current_data$category,
        test_group = test_group,
        n_randomizations = 999
      )
    
    observed_difference <-
      difference_of_medians(
        trait_values =
          current_data[[trait_column]],
        categories =
          current_data$category,
        test_group = test_group
      )
    
    results$observed_difference[i] <-
      observed_difference
    
    results$P[i] <-
      as.randtest(
        sim = simulated_differences,
        obs = observed_difference,
        alter = "two-sided"
      )$pvalue
  }
  
  results %>%
    group_by(trait) %>%
    mutate(
      Padj = p.adjust(
        P,
        method = "fdr"
      )
    ) %>%
    ungroup() %>%
    mutate(
      topography = recode(
        topography,
        !!!topography_labels
      )
    )
}

native_asteraceae_trait_tests <-
  run_single_trait_tests(
    topographic_data = topographic_data,
    test_group = "Aster_native"
  )

exotic_asteraceae_trait_tests <-
  run_single_trait_tests(
    topographic_data = topographic_data,
    test_group = "Aster_exotic"
  )

native_asteraceae_trait_tests %>%
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 3)
    )
  )

exotic_asteraceae_trait_tests %>%
  mutate(
    across(
      where(is.numeric),
      ~ round(.x, 3)
    )
  )

```


