---
title: "Manuscript figures and tables"
output: html_document
editor_options:
  chunk_output_type: console
---

## Purpose

This script reproduces manuscript figures and tables from the deposited vegetation and imputed trait datasets.

Figure 2 is presented as three two-dimensional trait-space plots rather than a single three-dimensional plot. Points are semi-transparent to reduce overplotting. Each trait pair is shown separately for the four topographic groups.

```{r setup}

library(tidyverse)
library(ggplot2)
library(patchwork)
library(readr)
library(ggpubr)
library(grid)

```

## Figure 1

```{r Fig 1: read-data}

analysis_unscaled <- read_csv(
  "analysis_unscaled.csv",
  show_col_types = FALSE
)

native_trait_tests <- read_csv(
  "native_asteraceae_trait_tests.csv",
  show_col_types = FALSE
)

exotic_trait_tests <- read_csv(
  "exotic_asteraceae_trait_tests.csv",
  show_col_types = FALSE
)

```

```{r Fig 1: single-trait}

# separate individual topographies
trait_upper_slope.ns <- analysis_unscaled %>%
  filter(upper_slope == 1)

trait_lower_slope.ns <- analysis_unscaled %>%
  filter(lower_slope == 1)

trait_ridge_summit.ns <- analysis_unscaled %>%
  filter(ridge_summit == 1)

trait_valley_flatland.ns <- analysis_unscaled %>%
  filter(valley_flatland == 1)

# combine the first group of topographies (Upper Slope, Lower Slope)
PlotData <- bind_rows(
  trait_upper_slope.ns %>% mutate(Topography = "Upper Slope"),
  trait_lower_slope.ns %>% mutate(Topography = "Lower Slope"),
  trait_ridge_summit.ns %>% mutate(Topography = "Ridge + Summit"),
  trait_valley_flatland.ns %>% mutate(Topography = "Valley + Flatland")
) %>%
  filter(category != "cultivar") %>%
  mutate(category = factor(category, levels = c("native", "Aster_native", "Aster_exotic")))

# log transform traits
PlotData <- PlotData %>%
  mutate(SLA_log = log10(SLA_mm2_mg.1),
         height_log = log10(height_m),
         seed_mass_log = log10(seed_dry_mass_mg))

# prepare p-values and significance labels for plotting
reference_trait_tests <- expand_grid(
  Topography = c(
    "Upper Slope",
    "Lower Slope",
    "Ridge + Summit",
    "Valley + Flatland"
  ),
  trait = c(
    "SLA_mm2_mg.1",
    "height_m",
    "seed_dry_mass_mg"
  )
) %>%
  mutate(
    category = "native",
    p = 1
  )

native_trait_plot_tests <- native_trait_tests %>%
  transmute(
    Topography = topography,
    trait = recode(
      trait,
      SLA = "SLA_mm2_mg.1",
      height = "height_m",
      seed_mass = "seed_dry_mass_mg"
    ),
    category = "Aster_native",
    p = P
  )

exotic_trait_plot_tests <- exotic_trait_tests %>%
  transmute(
    Topography = topography,
    trait = recode(
      trait,
      SLA = "SLA_mm2_mg.1",
      height = "height_m",
      seed_mass = "seed_dry_mass_mg"
    ),
    category = "Aster_exotic",
    p = P
  )

SLAP.PHAT.GERM <- bind_rows(
  reference_trait_tests,
  native_trait_plot_tests,
  exotic_trait_plot_tests
) %>%
  mutate(
    category = factor(
      category,
      levels = c(
        "native",
        "Aster_native",
        "Aster_exotic"
      )
    ),
    sig = case_when(
      p <= 0.001 ~ "***",
      p <= 0.01 ~ "**",
      p <= 0.05 ~ "*",
      TRUE ~ ""
    )
  )

# create site attribute object that contains the count of plots from PlotData by plant category and by topography
site.attr <- PlotData %>% 
  count(.,category,Topography) %>%
  mutate(n = paste0("n=",n))

# establish where on plots the p-value star indicators are displayed. three objects created, corresponding to each trait
loc.SLA <- PlotData %>% 
  summarise(., top.label = max(SLA_mm2_mg.1) + sd(SLA_mm2_mg.1), bottom.label = min(SLA_mm2_mg.1) - 0.25*sd(SLA_mm2_mg.1)) %>% 
  mutate(trait = "SLA_mm2_mg.1") %>% 
  dplyr::select(trait, top.label, bottom.label)

loc.PH <- PlotData %>% 
  summarise(., top.label = max(height_m) + sd(height_m), bottom.label = 0.05) %>% 
  mutate(trait = "height_m") %>% 
  dplyr::select(trait, top.label, bottom.label)

loc.GERM <- PlotData %>% 
  summarise(., top.label = max(seed_dry_mass_mg) + sd(seed_dry_mass_mg), bottom.label = 0.001) %>% 
  mutate(trait = "seed_dry_mass_mg") %>% 
  dplyr::select(trait, top.label, bottom.label)

# combine these three trait objects with label locations into one object loc(ations)
loc <- rbind(loc.SLA,loc.PH,loc.GERM)

# combine everything, including combinations of trait, plant category, topography and pvalue (SLAP.PHAT.GERM), sample size for each combination (site.attr), and the label location (loc)
SLAP.PHAT.GERM.full <- SLAP.PHAT.GERM %>%
  left_join(
    site.attr,
    by = c(
      "category",
      "Topography"
    )
  ) %>%
  left_join(
    loc,
    by = "trait"
  )

# make sure topography variable has the correct order of factors in the p-value object (SLAP.PHAT.GERM.full)
SLAP.PHAT.GERM.full$Topography <- factor(SLAP.PHAT.GERM.full$Topography, levels = c("Upper Slope","Lower Slope", "Ridge + Summit", "Valley + Flatland"))

#and in the PlotData object (PlotData)
PlotData$Topography <- factor(PlotData$Topography, levels = c("Upper Slope","Lower Slope", "Ridge + Summit", "Valley + Flatland"))

# colors for boxplots
box_colors <- c("native" = "mediumseagreen", "Aster_native" = "deepskyblue", "Aster_exotic" = "red2")

# define a function for plotting each trait
plot_trait <- function(trait, y_label, bottom = FALSE) {
  
  label.df <- SLAP.PHAT.GERM.full[which(SLAP.PHAT.GERM.full$trait==trait),]
  
  if(bottom == FALSE) {ggplot(PlotData, aes(x = category, y = get(trait), color = category)) +
    geom_jitter(aes(fill = NULL), width = 0.2, size = .25) +  # Add jittered points
    stat_boxplot(geom ='errorbar', alpha = 0.4, color = "black", linewidth = 0.5) + 
    geom_boxplot(aes(fill = category),outlier.shape = NA, alpha = 0.4, color = "black", linewidth = 0.5) +  # # remove outliers from boxplot
    facet_wrap(~ Topography, ncol = 2) +  #split by Topography
    geom_text(data = label.df, aes(label=n, y = top.label),size= 2 , color = "black", vjust = -0.5)+
    geom_text(data = label.df, aes(label=sig, y = bottom.label),size = 3, color = "black")+
    scale_color_manual(values = box_colors) +
    scale_fill_manual(values = box_colors) + 
    scale_y_continuous(trans = "log10", breaks = c(0.01, 0.1, 1, 10, 100, 1000), labels = c(0.01, 0.1, 1, 10, 100, 1000),expand = c(0.1, 0, 0.2, 0)) +
    scale_x_discrete(labels = c("Native\nHerbaceous","Native\nAsteraceae","Exotic\nAsteraceae"))+
    labs(x = "Category", y = y_label) +
    theme_bw() +
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          axis.title.y = element_text(size = 10),
          axis.title.x = element_blank(),
          strip.text.x = element_text(size = 10), #sets facet font size to axis size
          strip.background = element_rect(fill = "white"), #removes gray background from facet labels
          legend.position = "none")}
  
  else {ggplot(PlotData, aes(x = category, y = get(trait), color = category)) +
    geom_jitter(aes(fill = NULL), width = 0.2, size = .25) +  #add jittered points
    stat_boxplot(geom ='errorbar', alpha = 0.4, color = "black", linewidth = 0.5) + 
    geom_boxplot(aes(fill = category),outlier.shape = NA, alpha = 0.4, color = "black", linewidth = 0.5) +  # # remove outliers from boxplot
    facet_wrap(~ Topography, ncol = 2) +  #split by topography
    geom_text(data = label.df, aes(label=n, y = top.label),size=2, color = "black", vjust = -0.5)+
    geom_text(data = label.df, aes(label=sig, y = bottom.label),size=3, color = "black")+
    scale_color_manual(values = box_colors) +
    scale_fill_manual(values = box_colors) + 
    scale_y_continuous(trans = "log10", breaks = c(0.01, 0.1, 1, 10, 100, 1000), labels = c(0.01, 0.1, 1, 10, 100, 1000),expand = c(0.1, 0, 0.2, 0)) +
    scale_x_discrete(labels = c("Native\nHerbaceous","Native\nAsteraceae","Exotic\nAsteraceae"))+
    labs(x = "Category", y = y_label) +
    theme_bw() +
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          axis.title.y = element_text(size = 10),
          axis.title.x = element_text(size = 10),
          strip.text.x = element_text(size = 10), #sets facet font size to axis size
          strip.background = element_rect(fill = "white"), #removes gray background from facet labels
          legend.position = "none")}
}

# plot each trait
plot_SLA <- plot_trait("SLA_mm2_mg.1", expression(SLA~(mm^2~mg^-1)), bottom = FALSE)
plot_height <- plot_trait("height_m", expression(Height~(m)), bottom = FALSE)
plot_seed_mass <- plot_trait("seed_dry_mass_mg", expression(Seed~weight~(mg)), bottom = TRUE)

# horizontal stack: themes added to adjust the x axis labels, and either include x axis title for the center plot (plot_height.H) or exclude x axis title for right-most plot that had it in the vertical configuration (plot_seed_mass.H)

# each line adjusts specific things, and makes things the appropriate size.
plot_SLA.H <- plot_SLA+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size = 6),  #adjusts x axis text
        axis.text.y = element_text(size = 6), #adjusts y axis text
        axis.title.y = element_text(size = 8), #adjusts y axis title
        strip.text.x = element_text(size = 6)) #adjusts facet labels
plot_height.H <- plot_height+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size = 6), 
        axis.text.y = element_text(size = 6),  #adjusts y axis text
        axis.title.x = element_text(size = 8), #adjusts x axis title
        axis.title.y = element_text(size = 8), #adjusts y axis title
        strip.text.x = element_text(size = 6)) #adjusts facet labels
plot_seed_mass.H <- plot_seed_mass + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size = 6), 
        axis.text.y = element_text(size = 6),  #adjusts y axis text
        axis.title.x = element_blank(), #removes X axis title
        axis.title.y = element_text(size = 8), #adjusts y axis title
        strip.text.x = element_text(size = 6)) #adjusts facet labels

# create horizontal final figure 1
figure1 <- ggarrange(
  plot_SLA.H,
  plot_height.H,
  plot_seed_mass.H,
  ncol = 3,
  align = "hv"
)

figure1


```

## Figure 2

```{r Fig 2: read-data}

analysis_scaled <- read_csv(
  "analysis_scaled.csv",
  show_col_types = FALSE
)

```

## Prepare data for Figure 2

The scaled dataset contains log10-transformed and standardized trait values generated in 02_trait_imputation.Rmd.

```{r Fig 2: prepare-figure-data}

figure_data <- analysis_scaled %>%
  select(
    species,
    category,
    height_m,
    SLA_mm2_mg.1,
    seed_dry_mass_mg,
    upper_slope,
    lower_slope,
    ridge_summit,
    valley_flatland
  ) %>%
  pivot_longer(
    cols = c(
      upper_slope,
      lower_slope,
      ridge_summit,
      valley_flatland
    ),
    names_to = "topography",
    values_to = "presence"
  ) %>%
  filter(
    presence == 1
  ) %>%
  mutate(
    category = factor(
      category,
      levels = c(
        "native",
        "Aster_native",
        "Aster_exotic"
      )
    ),
    topography = factor(
      topography,
      levels = c(
        "upper_slope",
        "lower_slope",
        "ridge_summit",
        "valley_flatland"
      ),
      labels = c(
        "Upper Slope",
        "Lower Slope",
        "Ridge + Summit",
        "Valley + Flatland"
      )
    )
  ) %>%
  arrange(
    topography,
    category
  )

table(
  figure_data$topography,
  figure_data$category
)

```

## Figure 2: 2D projections of trait space

```{r figure-2}

category_colors <- c(
  "native" = "mediumseagreen",
  "Aster_native" = "deepskyblue",
  "Aster_exotic" = "red2"
)

category_labels <- c(
  "native" = "Native non-Asteraceae",
  "Aster_native" = "Native Asteraceae",
  "Aster_exotic" = "Exotic Asteraceae"
)

trait_plot_theme <- theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 9),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 8),
    legend.title = element_blank()
  )

height_sla_plot <- ggplot(
  figure_data,
  aes(
    x = height_m,
    y = SLA_mm2_mg.1,
    color = category
  )
) +
  geom_point(
    alpha = 0.5,
    size = 1.8
  ) +
  facet_wrap(
    ~ topography,
    ncol = 2
  ) +
  scale_color_manual(
    values = category_colors,
    labels = category_labels
  ) +
  labs(
    x = expression(Standardized~height),
    y = expression(Standardized~SLA),
    title = "A Height vs. SLA"
  ) +
  trait_plot_theme +
  theme(
    legend.position = "none"
  )

height_seed_plot <- ggplot(
  figure_data,
  aes(
    x = height_m,
    y = seed_dry_mass_mg,
    color = category
  )
) +
  geom_point(
    alpha = 0.5,
    size = 1.8
  ) +
  facet_wrap(
    ~ topography,
    ncol = 2
  ) +
  scale_color_manual(
    values = category_colors,
    labels = category_labels
  ) +
  labs(
    x = expression(Standardized~height),
    y = expression(Standardized~seed~mass),
    title = "B Height vs. Seed mass"
  ) +
  trait_plot_theme +
  theme(
    legend.position = "none"
  )

sla_seed_plot <- ggplot(
  figure_data,
  aes(
    x = SLA_mm2_mg.1,
    y = seed_dry_mass_mg,
    color = category
  )
) +
  geom_point(
    alpha = 0.5,
    size = 1.8
  ) +
  facet_wrap(
    ~ topography,
    ncol = 2
  ) +
  scale_color_manual(
    values = category_colors,
    labels = category_labels
  ) +
  labs(
    x = expression(Standardized~SLA),
    y = expression(Standardized~seed~mass),
    title = "C SLA vs. Seed mass"
  ) +
  trait_plot_theme +
  theme(
    legend.position = "bottom"
  )

figure_2 <- (
  height_sla_plot /
  height_seed_plot /
  sla_seed_plot
) +
  plot_layout(
    guides = "collect"
  ) &
  theme(
    legend.position = "bottom",
  )

figure_2

```



